---
output: html_document
editor_options: 
  chunk_output_type: inline
---



IDENTIFICATION

# (S)ARIMA(p, d, q)(P, D, Q)_s
## (Seasonal) Auto-Regressive Integrated Moving Average
### (p, d, q) - Regular dependence
### (P, D, Q) - Seasonal dependence

ESTIMATION

arima(lnserie, order = c(p, d, q), seasonal = list(order = c(P, D, Q), period = s))



## AirBCN

```{r}
serie=window(ts(read.csv2("airbcn.dat",header=FALSE)/1000,start=1990,freq=12),start=1990)
print(round(serie,0))
```

```{r}
plot(serie,main="International Airlines Passengers in BCN")
abline(v=1990:2020,col=4,lty=3)
```

```{r}
lnserie=log(serie)
plot(lnserie)
var(lnserie)
```

```{r}
d12lnserie=diff(lnserie,12)
plot(d12lnserie)
abline(h=0)
var(d12lnserie)
```

```{r}
d1d12lnserie=diff(d12lnserie,1)
plot(d1d12lnserie)
abline(h=0)
var(d1d12lnserie)

var(diff(d1d12lnserie)) # The variance has increased - STOP
```



```{r}
################# Validation #################################
validation=function(model){
  s=frequency(get(model$series))
  resi=model$residuals
  par(mfrow=c(2,2),mar=c(3,3,3,3))
  #Residuals plot
  plot(resi,main="Residuals")
  abline(h=0)
  abline(h=c(-3*sd(resi),3*sd(resi)),lty=3,col=4)
  #Square Root of absolute values of residuals (Homocedasticity)
  scatter.smooth(sqrt(abs(resi)),main="Square Root of Absolute residuals",
                 lpars=list(col=2))
  
  #Normal plot of residuals
  qqnorm(resi)
  qqline(resi,col=2,lwd=2)
  
  ##Histogram of residuals with normal curve
  hist(resi,breaks=20,freq=FALSE)
  curve(dnorm(x,mean=mean(resi),sd=sd(resi)),col=2,add=T)
  
  
  #ACF & PACF of residuals
  par(mfrow=c(1,2))
  acf(resi,ylim=c(-1,1),lag.max=60,col=c(2,rep(1,s-1)),lwd=2)
  pacf(resi,ylim=c(-1,1),lag.max=60,col=c(rep(1,s-1),2),lwd=2)
  par(mfrow=c(1,1))
  
  #Ljung-Box p-values
  par(mar=c(2,2,1,1))
  tsdiag(model,gof.lag=7*s)
  cat("\n--------------------------------------------------------------------\n")
  print(model)
  
  #Stationary and Invertible
  cat("\nModul of AR Characteristic polynomial Roots: ", 
      Mod(polyroot(c(1,-model$model$phi))),"\n")
  cat("\nModul of MA Characteristic polynomial Roots: ",
      Mod(polyroot(c(1,model$model$theta))),"\n")
  
  suppressMessages(require(forecast,quietly=TRUE,warn.conflicts=FALSE))
  plot(model)
  
  #Model expressed as an MA infinity (psi-weights)
  psis=ARMAtoMA(ar=model$model$phi,ma=model$model$theta,lag.max=36)
  names(psis)=paste("psi",1:36)
  cat("\nPsi-weights (MA(inf))\n")
  cat("\n--------------------\n")
  print(psis[1:24])
  
  #Model expressed as an AR infinity (pi-weights)
  pis=-ARMAtoMA(ar=-model$model$theta,ma=-model$model$phi,lag.max=36)
  names(pis)=paste("pi",1:36)
  cat("\nPi-weights (AR(inf))\n")
  cat("\n--------------------\n")
  print(pis[1:24])
   
  cat("\nDescriptive Statistics for the Residuals\n")
  cat("\n----------------------------------------\n") 
  
  suppressMessages(require(fBasics,quietly=TRUE,warn.conflicts=FALSE))
  ##Anderson-Darling test
  print(basicStats(resi))
  
  ## Add here complementary tests (use with caution!)
  ##---------------------------------------------------------
  cat("\nNormality Tests\n")
  cat("\n--------------------\n")
 
  ##Shapiro-Wilks Normality test
  print(shapiro.test(resi))

  suppressMessages(require(nortest,quietly=TRUE,warn.conflicts=FALSE))
  ##Anderson-Darling test
  print(ad.test(resi))
  
  suppressMessages(require(tseries,quietly=TRUE,warn.conflicts=FALSE))
  ##Jarque-Bera test
  print(jarque.bera.test(resi))
  
  cat("\nHomoscedasticity Test\n")
  cat("\n--------------------\n")
  suppressMessages(require(lmtest,quietly=TRUE,warn.conflicts=FALSE))
  ##Breusch-Pagan test
  obs=get(model$series)
  print(bptest(resi~I(obs-resi)))
  
  cat("\nIndependence Tests\n")
  cat("\n--------------------\n")
  
  ##Durbin-Watson test
  print(dwtest(resi~I(1:length(resi))))
  
  ##Ljung-Box test
  cat("\nLjung-Box test\n")
  print(t(apply(matrix(c(1:4,(1:4)*s)),1,function(el) {
    te=Box.test(resi,type="Ljung-Box",lag=el)
    c(lag=(te$parameter),statistic=te$statistic[[1]],p.value=te$p.value)})))
  
}
################# Fi Validation #################################
```

```{r}
par(mfrow=c(1,2))
acf(d1d12lnserie,ylim=c(-1,1),col=c(2,rep(1,11)),lwd=2,lag.max=72)
pacf(d1d12lnserie,ylim=c(-1,1),col=c(rep(1,11),2),lwd=2,lag.max=72)
par(mfrow=c(1,1))
```

# Identification of the model

Two possibilities:
ARIMA(0, 1, 1)(2, 1, 0)_12
ARIMA(2, 1, 0)(2, 1, 0)_12

d: the power of the transformation, usually equals to 1

```{r}
(mod1=arima(lnserie,order=c(0,1,1),seasonal=list(order=c(2,1,0),period=12)))
```

# Estimation of the model

(1 - Fi_1 * B^12 -  Fi_2 * B24) * ((1 - B) * (1 - B^12) * log(X_t)) = (1 + Theta_1 * B) * Z_t
[          AR(2)_12           ]   [              W_t              ]   [     MA(1)     ]

(1 + 0.63 * B^12 + 0.40 * B^24) * ((1 - B) * (1 - B^12) * log(X_t)) = (1 - 0.37 * B) * Z_t


# Residual analysis

Z_t = X_t - X^{~}_{t|t-1}
where X_t is observation and X^{~}_{t|t-1} is prediction of X_t having X_1, ..., X_{t-1}

-> Constant Variance -> Homocedapticity
-> Normality (Normal Q-Q Plot better than the Histogram plot)
-> Independence (All the points on the 'p value for Ljung-Box statistic' plot have to be above the line)

To validate the model successfully, each of the 3 things from above have to be checked!

It is expected that 99,7% of all the data is inside the (3 sigma : -3 sigma) interval.

```{r}
validation(mod1)
```

```{r}
(mod2=arima(lnserie,order=c(2,1,0),seasonal=list(order=c(2,1,0),period=12)))
```

```{r}
validation(mod2)
```

```{r}
(mod3=arima(lnserie,order=c(0,1,1),seasonal=list(order=c(0,1,1),period=12)))

```

```{r}
validation(mod3)
```

```{r}
ultim=c(2018,12)
pdq=c(0,1,1)
PDQ=c(2,1,0)

serie2=window(serie,end=ultim)
lnserie2=log(serie2)
serie1=window(serie,end=ultim+c(1,0))
lnserie1=log(serie1)

(modA=arima(lnserie1,order=pdq,seasonal=list(order=PDQ,period=12)))
(modB=arima(lnserie2,order=pdq,seasonal=list(order=PDQ,period=12)))
```


```{r}
pred=predict(modB,n.ahead=12)
pr<-ts(c(tail(lnserie2,1),pred$pred),start=ultim,freq=12)
se<-ts(c(0,pred$se),start=ultim,freq=12)

#Intervals
tl<-ts(exp(pr-1.96*se),start=ultim,freq=12)
tu<-ts(exp(pr+1.96*se),start=ultim,freq=12)
pr<-ts(exp(pr),start=ultim,freq=12)


ts.plot(serie,tl,tu,pr,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=ultim[1]+c(-2,+2),type="o",main=paste("Model ARIMA(",paste(pdq,collapse=","),")(",paste(PDQ,collapse=","),")12",sep=""))
abline(v=(ultim[1]-2):(ultim[1]+2),lty=3,col=4)
```

```{r}
obs=window(serie,start=ultim+c(0,1))
pr=window(pr,start=ultim+c(0,1))
ts(data.frame(LowLim=tl[-1],Predic=pr,UpperLim=tu[-1],Observ=obs,Error=obs-pr,PercentError=(obs-pr)/obs),start=ultim+c(0,1),freq=12)
mod.RMSE1=sqrt(sum((obs-pr)^2)/12)
mod.MAE1=sum(abs(obs-pr))/12
mod.RMSPE1=sqrt(sum(((obs-pr)/obs)^2)/12)
mod.MAPE1=sum(abs(obs-pr)/obs)/12

data.frame("RMSE"=mod.RMSE1,"MAE"=mod.MAE1,"RMSPE"=mod.RMSPE1,"MAPE"=mod.MAPE1)

mCI1=mean(tu-tl)

cat("\nMean Length CI: ",mCI1)
```

```{r}
pred=predict(modA,n.ahead=12)
pr<-ts(c(tail(lnserie1,1),pred$pred),start=ultim+c(1,0),freq=12)
se<-ts(c(0,pred$se),start=ultim+c(1,0),freq=12)

tl1<-ts(exp(pr-1.96*se),start=ultim+c(1,0),freq=12)
tu1<-ts(exp(pr+1.96*se),start=ultim+c(1,0),freq=12)
pr1<-ts(exp(pr),start=ultim+c(1,0),freq=12)

ts.plot(serie,tl1,tu1,pr1,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=c(ultim[1]-2,ultim[1]+3),type="o",main=paste("Model ARIMA(",paste(pdq,collapse=","),")(",paste(PDQ,collapse=","),")12",sep=""))
abline(v=(ultim[1]-2):(ultim[1]+3),lty=3,col=4)
```

```{r}
(previs1=window(cbind(tl1,pr1,tu1),start=ultim+c(1,0)))
```

```{r}
ultim=c(2018,12)
pdq=c(2,1,0)
PDQ=c(2,1,0)

serie2=window(serie,end=ultim)
lnserie2=log(serie2)
serie1=window(serie,end=ultim+c(1,0))
lnserie1=log(serie1)

(modA=arima(lnserie1,order=pdq,seasonal=list(order=PDQ,period=12)))
(modB=arima(lnserie2,order=pdq,seasonal=list(order=PDQ,period=12)))
```

```{r}
pred=predict(modB,n.ahead=12)
pr<-ts(c(tail(lnserie2,1),pred$pred),start=ultim,freq=12)
se<-ts(c(0,pred$se),start=ultim,freq=12)

#Intervals
tl<-ts(exp(pr-1.96*se),start=ultim,freq=12)
tu<-ts(exp(pr+1.96*se),start=ultim,freq=12)
pr<-ts(exp(pr),start=ultim,freq=12)


ts.plot(serie,tl,tu,pr,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=ultim[1]+c(-2,+2),type="o",main=paste("Model ARIMA(",paste(pdq,collapse=","),")(",paste(PDQ,collapse=","),")12",sep=""))
abline(v=(ultim[1]-2):(ultim[1]+2),lty=3,col=4)
```

```{r}
obs=window(serie,start=ultim+c(0,1))
pr=window(pr,start=ultim+c(0,1))
ts(data.frame(LowLim=tl[-1],Predic=pr,UpperLim=tu[-1],Observ=obs,Error=obs-pr,PercentError=(obs-pr)/obs),start=ultim+c(0,1),freq=12)
mod.RMSE2=sqrt(sum((obs-pr)^2)/12)
mod.MAE2=sum(abs(obs-pr))/12
mod.RMSPE2=sqrt(sum(((obs-pr)/obs)^2)/12)
mod.MAPE2=sum(abs(obs-pr)/obs)/12

data.frame("RMSE"=mod.RMSE2,"MAE"=mod.MAE2,"RMSPE"=mod.RMSPE2,"MAPE"=mod.MAPE2)

mCI2=mean(tu-tl)

cat("\nMean Length CI: ",mCI1)
```

```{r}
pred=predict(modA,n.ahead=12)
pr<-ts(c(tail(lnserie1,1),pred$pred),start=ultim+c(1,0),freq=12)
se<-ts(c(0,pred$se),start=ultim+c(1,0),freq=12)

tl1<-ts(exp(pr-1.96*se),start=ultim+c(1,0),freq=12)
tu1<-ts(exp(pr+1.96*se),start=ultim+c(1,0),freq=12)
pr1<-ts(exp(pr),start=ultim+c(1,0),freq=12)

ts.plot(serie,tl1,tu1,pr1,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=c(ultim[1]-2,ultim[1]+3),type="o",main=paste("Model ARIMA(",paste(pdq,collapse=","),")(",paste(PDQ,collapse=","),")12",sep=""))
abline(v=(ultim[1]-2):(ultim[1]+3),lty=3,col=4)
```

```{r}
(previs2=window(cbind(tl1,pr1,tu1),start=ultim+c(1,0)))
```

```{r}
ultim=c(2018,12)
pdq=c(0,1,1)
PDQ=c(0,1,1)

serie2=window(serie,end=ultim)
lnserie2=log(serie2)
serie1=window(serie,end=ultim+c(1,0))
lnserie1=log(serie1)

(modA=arima(lnserie1,order=pdq,seasonal=list(order=PDQ,period=12)))
(modB=arima(lnserie2,order=pdq,seasonal=list(order=PDQ,period=12)))
```

```{r}
pred=predict(modB,n.ahead=12)
pr<-ts(c(tail(lnserie2,1),pred$pred),start=ultim,freq=12)
se<-ts(c(0,pred$se),start=ultim,freq=12)

#Intervals
tl<-ts(exp(pr-1.96*se),start=ultim,freq=12)
tu<-ts(exp(pr+1.96*se),start=ultim,freq=12)
pr<-ts(exp(pr),start=ultim,freq=12)


ts.plot(serie,tl,tu,pr,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=ultim[1]+c(-2,+2),type="o",main=paste("Model ARIMA(",paste(pdq,collapse=","),")(",paste(PDQ,collapse=","),")12",sep=""))
abline(v=(ultim[1]-2):(ultim[1]+2),lty=3,col=4)
```

```{r}
obs=window(serie,start=ultim+c(0,1))
pr=window(pr,start=ultim+c(0,1))
ts(data.frame(LowLim=tl[-1],Predic=pr,UpperLim=tu[-1],Observ=obs,Error=obs-pr,PercentError=(obs-pr)/obs),start=ultim+c(0,1),freq=12)
mod.RMSE3=sqrt(sum((obs-pr)^2)/12)
mod.MAE3=sum(abs(obs-pr))/12
mod.RMSPE3=sqrt(sum(((obs-pr)/obs)^2)/12)
mod.MAPE3=sum(abs(obs-pr)/obs)/12

data.frame("RMSE"=mod.RMSE2,"MAE"=mod.MAE2,"RMSPE"=mod.RMSPE2,"MAPE"=mod.MAPE2)

mCI3=mean(tu-tl)

cat("\nMean Length CI: ",mCI1)
```

```{r}
pred=predict(modA,n.ahead=12)
pr<-ts(c(tail(lnserie1,1),pred$pred),start=ultim+c(1,0),freq=12)
se<-ts(c(0,pred$se),start=ultim+c(1,0),freq=12)

tl1<-ts(exp(pr-1.96*se),start=ultim+c(1,0),freq=12)
tu1<-ts(exp(pr+1.96*se),start=ultim+c(1,0),freq=12)
pr1<-ts(exp(pr),start=ultim+c(1,0),freq=12)

ts.plot(serie,tl1,tu1,pr1,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=c(ultim[1]-2,ultim[1]+3),type="o",main=paste("Model ARIMA(",paste(pdq,collapse=","),")(",paste(PDQ,collapse=","),")12",sep=""))
abline(v=(ultim[1]-2):(ultim[1]+3),lty=3,col=4)
```

```{r}
(previs3=window(cbind(tl1,pr1,tu1),start=ultim+c(1,0)))
```
```{r}

resul=data.frame(
  par=c(length(coef(mod1)),length(coef(mod2)),length(coef(mod3))),
  Sigma2Z=c(mod1$sigma2,mod2$sigma2,mod3$sigma2),
  AIC=c(AIC(mod1),AIC(mod2),AIC(mod3)),
  BIC=c(BIC(mod1),BIC(mod2),BIC(mod3)),
   RMSE=c(mod.RMSE1,mod.RMSE2,mod.RMSE3),
  MAE=c(mod.MAE1,mod.MAE2,mod.MAE3),
  RMSPE=c(mod.RMSPE1,mod.RMSPE2,mod.RMSPE3),
  MAPE=c(mod.MAPE1,mod.MAPE2,mod.MAPE3),
  meanLength=c(mCI1,mCI2,mCI3)
  )

row.names(resul)=c("ARIMA(0,1,1)(2,1,0)12","ARIMA(2,1,0)(2,1,0)12","ARIMA(0,1,1)(0,1,1)12")



resul


```

